---
title: "Atlantic 16S Analysis – Sea4Blue Transect"
author: "Your Name"
format: html
editor: visual
---

# Introduction

This report presents the spatial and environmental characterization of the Atlantic Sea4Blue 16S dataset.\
Sampling stations are mapped across Longhurst biogeographical provinces and overlaid with annual Sea Surface Temperature (SST) derived from MODIS satellite data.

------------------------------------------------------------------------
# Load Required Libraries


```{r}
#| message: false
#| warning: false
library(ggforce)
library(Matrix)
library(foreach)
library(pander)
library(doParallel)
library(ape)
library(ggpubr)
library(ggrepel)
library(ggplot2)
library(sf)
library(rnaturalearth)
library(raster)
library(ncdf4)
library(geosphere)
library(tibble)
library(oceanmap)
library(RColorBrewer)
library(viridis)
library(scales)
library(dplyr)
library(umap)
library(phyloseq)
library(microbiome)
library(MiscMetabar)
library(tidyverse)
library(vegan)
library(picante)
library(Biostrings)
library(ape)
```

Load Phyloseq Object

```{r}
path <- "/media/shared1/atlantic_sea4blue/16S_Illumina/Lapo_18May/"

ps <- readRDS(paste0(path, "ps.rds"))

# Inspect OTU table
head(t(otu_table(ps)))
```

# Plot Sea4Blue transect

Extract Sample Coordinates

```{r}
coord.df <- ps %>%
  subset_samples(!(sample.description %in% c("Laposnegative", 
                                            "Lapospositive"))) %>%
  meta() %>%
  dplyr::select(sample.description, Latitude, Longitude)

colnames(coord.df) <- c("id", "lat", "lon")

head(coord.df)

points_sf <- st_as_sf(coord.df,
                      coords = c("lon", "lat"),
                      crs = 4326)
```

Load Longhurst Provinces

```{r}
longhurst <- st_read(paste0(path, "longhurst_v4_2010/Longhurst_world_v4_2010.shp"),
                     quiet = TRUE)

longhurst <- st_make_valid(longhurst)
longhurst <- st_transform(longhurst, 4326)

result <- st_join(points_sf,
                  longhurst["ProvCode"],
                  join = st_intersects)

head(as.data.frame(result))

```

Plot Longhurst Provinces and Sampling Stations

```{r}
world <- ne_countries(scale = "medium", returnclass = "sf")

province_used <- unique(result$ProvCode)

longhurst_subset <- longhurst %>%
  filter(ProvCode %in% province_used)

ggplot() +
  geom_sf(data = world, fill = "#D9D9D9",
          color = "gray70", size = 0.3) +
  geom_sf(data = longhurst_subset,
          aes(fill = ProvCode),
          color = NA, alpha = 0.4) +
  geom_sf(data = result,
          color = "black", size = 2) +
  geom_sf_text(data = result,
               aes(label = id),
               size = 3, nudge_y = 1,
               check_overlap = TRUE) +
  scale_fill_viridis_d(name = "Longhurst Province") +
  coord_sf(crs = 4326,
           xlim = c(-90, 10),
           ylim = c(25, 45)) +
  theme_minimal() +
  labs(title = "Longhurst Provinces and Sample Points",
       x = "Longitude",
       y = "Latitude")
```

```{r}
sample_data(ps)$longhurst_provinces <-
  result$ProvCode[
    match(sample_data(ps)$sample.description,
          result$id)
  ]
```

Download Annual Sea Surface Temperature (MODIS) externally using terminal:

```         
wget --auth-no-challenge --load-cookies ~/.urs_cookies \
--save-cookies ~/.urs_cookies \
"https://oceandata.sci.gsfc.nasa.gov/ob/getfile/AQUA_MODIS.20220101_20221231.L3m.YR.SST.sst.4km.nc"
```

------------------------------------------------------------------------

```{r}

crop.vals_Atl <- c(lat = c(-90,10), 
                   lon = c(25, 45))

SST_Atl <- ncdf4::nc_open(paste0(path,"AQUA_MODIS.20220101_20221231.L3m.YR.SST.sst.4km.nc")) %>%
  oceanmap::nc2raster(., "sst") %>%
  raster::flip(., "y") %>%
  raster::crop(., raster::extent(crop.vals_Atl))

SST_df <- as.data.frame(SST_Atl, xy = TRUE)
names(SST_df)[3] <- "sst" 
points_sf$longhurst_provinces <- as.data.frame(result)$ProvCode
```

#Combined Map: SST + Longhurst Provinces

```{r}
ggplot() +
  geom_raster(data = SST_df,
              aes(x = x, y = y, fill = sst)) +
  scale_fill_gradientn(
    colors = rev(brewer.pal(11, "RdBu")),
    name = "Annual SST (°C)",
    limits = c(0, 35),
    na.value = "transparent"
  ) +
  geom_sf(data = world,
          fill = "#D9D9D9",
          color = "gray70",
          size = 0.3) +
  geom_sf(data = longhurst_subset,
          color = "gray60",
          fill = NA,
          size = 0.4) +
  geom_sf(data = points_sf,
          aes(color = longhurst_provinces),
          size = 3) +
  scale_color_viridis_d(name = "Longhurst Province") +
  geom_sf_text(data = points_sf,
               aes(label = id),
               size = 3,
               nudge_y = 1,
               check_overlap = TRUE) +
  coord_sf(crs = 4326,
           xlim = c(-90, 10),
           ylim = c(25, 45),
           expand = FALSE) +
  theme_minimal() +
  labs(title = "Sea Surface Temperature and Sample Points",
       subtitle = "MODIS Annual SST + Longhurst Provinces",
       x = "Longitude",
       y = "Latitude")
```
# importo i dati scaricati in reticulate_var_amb_giuste.R
```{r}
rm(list = ls())

path <- "/media/shared1/atlantic_sea4blue/16S_Illumina/Lapo_18May/"

all_data<-read.csv(paste0(path,"dati_scaricati_copernicus.csv"))
ps <- readRDS(paste0(path,"ps.rds"))
otu_table(ps) <- otu_table(t(otu_table(ps)), taxa_are_rows = TRUE)
head(tax_table(ps))
phyloseq_obj<-ps
```


pre treatment data

```{r}

#rimuovo i contaminanti
grep(pattern = "Mitochondria", tax_table(phyloseq_obj)) 
grep(pattern = "Chloroplast", tax_table(phyloseq_obj)) 
phyloseq_obj <- phyloseq_obj %>% subset_taxa( Family!= "Mitochondria" | is.na(Family) & Class!="Chloroplast" | is.na(Class) ) 
phyloseq_obj <- subset_taxa(phyloseq_obj, (tax_table(phyloseq_obj)[,"Order"]!="Chloroplast") | is.na(tax_table(phyloseq_obj)[,"Order"]))
phyloseq_obj = subset_samples(phyloseq_obj,!( Nominativo.campione.=="Positive" | Nominativo.campione.=="Negative"))
sample_names(phyloseq_obj)<-sample_data(phyloseq_obj)$Nominativo.campione.

merged_metadata <- merge(sample_data(phyloseq_obj)%>% meta(), all_data, 
                         by = "Nominativo.campione.", 
                         all.x = TRUE)
rownames(merged_metadata) <- merged_metadata$Nominativo.campione.
sample_data(phyloseq_obj) <- merged_metadata
sample_data(phyloseq_obj)
phyloseq_obj<-clean_pq(phyloseq_obj)
```

# do folders and save into requiered files for subsequent analysis

```{r}
#dir.create("Data", recursive = TRUE)
#dir.create("Data/Meta_Data/Prok", recursive = TRUE)
#dir.create("Data/Count_Data/Processed/Prok", recursive = TRUE)
#dir.create("Data/Count_Data/Fasta/Prok", recursive = TRUE)
#dir.create("Data/Trees/Prok", recursive = TRUE)
#dir.create("Data/Processed/Prok", recursive = TRUE)
#dir.create("Data/Taxonomy", recursive = TRUE)

#metadata
#sample_data(phyloseq_obj) %>%
#   as_tibble() %>%
#  rename(Sample_ID = Nominativo.campione.) %>%
#  mutate(Sample_ID = as.character(Sample_ID)) %>%
#  relocate(Sample_ID, .before = 1)%>%
#  write_tsv("Data/Meta_Data/Prok/Meta_Data.tsv")
#asv table
#otu<-otu_table(phyloseq_obj) %>%
#  as.data.frame() %>%
#  rownames_to_column("OTU_ID") 
#  write_tsv(otu,"Data/Count_Data/Processed/Prok/Full_Prok_Count.tsv")  
#  write_tsv(otu,"Data/Processed/Prok/Full_Prok_Count.tsv")

#   Sequenze FASTA
#writeXStringSet(refseq(phyloseq_obj), "Data/Count_Data/Fasta/Prok/Full_Prok_Sequences.fasta")

# 4.4 Albero filogenetico
#write.tree(phy_tree(phyloseq_obj), "Data/Trees/Prok/Prok_Combined.tree")

# 4.5 Tassonomia
#tax_table(phyloseq_obj) %>%
#  as.data.frame() %>%
#  rownames_to_column("OTU_ID") %>%
#  write_tsv("Data/Taxonomy/Prok_Taxonomy.tsv")


```


Normalization
```{r}
doubleton <- genefilter_sample(phyloseq_obj, filterfun_sample(function(x) x > 1), A=1)
doubleton <- prune_taxa(doubleton, phyloseq_obj) 

# transforming 
data.metagenomeSeq = phyloseq_to_metagenomeSeq(doubleton)

p = cumNormStat(data.metagenomeSeq) #default is 0.5
data.cumnorm = cumNorm(data.metagenomeSeq, p=p)
#data.cumnorm
data.CSS = MRcounts(data.cumnorm, norm=TRUE, log=TRUE)
head(data.CSS)
dim(data.CSS)  # make sure the data are in a correct formal: number of samples in rows
phyloseq_obj_css <- phyloseq_obj
otu_table(phyloseq_obj_css) <- otu_table(data.CSS, taxa_are_rows = T)
sample_names(phyloseq_obj_css)<-sample_data(phyloseq_obj_css)$Nominativo.campione.
phyloseq_obj_css

```

# alpha diversity

```{r}
physeq_pa <- transform_sample_counts(phyloseq_obj_css, function(x) ifelse(x > 0, 1, 0))
head(otu_table(physeq_pa))

```
```{r}
sample_variables(physeq_pa)
```


```{r}
p1<-plot_richness(physeq_pa,x="Longitude.x",color = "Zona",measures=c("Observed"))
newSTorder = c( "ANW","ANC","ANE")
p1$data$Zona<- as.character(p1$data$Zona)
p1$data$Zona <- factor(p1$data$Zona, levels=newSTorder) 
ggplot(p1$data,aes(Longitude.x,value))+
  theme_bw()+theme(plot.title = element_text(hjust = 0.5))+
  geom_point(aes(colour = Zona))+
  scale_color_manual(values=c("#000000","#4280fc", "#ffb452","#f7170a"))+
  geom_ysideboxplot(aes(x=Zona,y=value,colour = Zona), orientation = "x") +
  theme(        ggside.panel.scale.y = .4)+scale_ysidex_discrete()+
  geom_smooth(aes(colour = variable ),size=1.5)+ ylab('Richness') 
```



```{r}

ggboxplot(p1$data, x = "Zona", y = "value",
               color = "Zona", palette = "jco",
               add = "jitter")+ stat_compare_means()   #0.06
```

# BETA DIVERSITY 
## braycurties
```{r}
otu.ord <- ordinate(physeq = phyloseq_obj_css, "PCoA", distance = "bray")

plot_ordination(phyloseq_obj_css, otu.ord,  color="Zona",axes =c(1,2))+
  scale_color_manual(values = c("#4280fc", "#ffb452","#f7170a"))+ geom_point(size=3)+ 
  geom_text_repel(aes(label=Giorno.di.navigazione.),max.overlaps = Inf, show.legend = FALSE)+
  geom_mark_ellipse(aes(color = Zona), show.legend = FALSE)+ theme_void()+theme_bw() + 
  geom_xsidedensity(aes(y=stat(density),fill=Zona), alpha = 0.5, show.legend = FALSE) +
  geom_ysidedensity(aes(x=stat(density),fill=Zona), alpha = 0.5, show.legend = FALSE) +
 scale_xsidey_continuous(breaks = NULL, labels = "", expand = expansion(c(0,.1))) +
  scale_ysidex_continuous(breaks = NULL, labels = "", expand = expansion(c(0,.1))) +
  ggside::theme_ggside_void()  +
  scale_fill_manual(values = c("#4280fc", "#ffb452","#f7170a"))+ggtitle("bray-Curtis") #+scale_ysidex_discrete()+
```


```{r}
set.seed(231)
bray <- phyloseq::distance(phyloseq_obj_css, method = "bray")
sampledf <- data.frame(sample_data(phyloseq_obj_css))
head(sampledf)
```

```{r}
adonis2(bray ~ Zona+Giorno.di.navigazione.+Latitude.x+Longitude.x+thetao, data = sampledf, by = "margin",permutations = 9999)
```

## wunifrac


```{r}
otu.ord <- ordinate(physeq = phyloseq_obj_css, "PCoA", distance = "wunifrac")

sample_data(phyloseq_obj_css)$Zona <-
  factor(sample_data(phyloseq_obj_css)$Zona)
#plot ordination

plot_ordination(phyloseq_obj_css, otu.ord,  color="Zona",axes =c(1,2))+
  scale_color_manual(values = c("#4280fc", "#ffb452","#f7170a"))+ geom_point(size=3)+ 
  geom_text_repel(aes(label=Giorno.di.navigazione.),max.overlaps = Inf, show.legend = FALSE)+
  geom_mark_ellipse(aes(color = Zona), show.legend = FALSE)+ theme_void()+theme_bw() + 
  geom_xsidedensity(aes(y=stat(density),fill=Zona), alpha = 0.5, show.legend = FALSE) +
  geom_ysidedensity(aes(x=stat(density),fill=Zona), alpha = 0.5, show.legend = FALSE) +
  scale_xsidey_continuous(breaks = NULL, labels = "", expand = expansion(c(0,.1))) +
  scale_ysidex_continuous(breaks = NULL, labels = "", expand = expansion(c(0,.1))) +
  ggside::theme_ggside_void()  +
  scale_fill_manual(values = c("#4280fc", "#ffb452","#f7170a"))+ggtitle("wunifrac")
```

```{r}
set.seed(1)

 uniF <- phyloseq::distance(phyloseq_obj_css, method = "wunifrac")
 sampledf <- data.frame(sample_data(phyloseq_obj_css))

 
adonis2(uniF ~ Zona+Giorno.di.navigazione.+thetao +Longitude.x, data = sampledf,by = "margin")
```

# distance decay

calcolare le matrici di distanza
```{r}
sample_data(phyloseq_obj_css)$Zona <-
  factor(sample_data(phyloseq_obj_css)$Zona)

sample_data(phyloseq_obj_css)$Latitude.y  <- as.numeric(sample_data(phyloseq_obj_css)$Latitude.y)
sample_data(phyloseq_obj_css)$Longitude.y <- as.numeric(sample_data(phyloseq_obj_css)$Longitude.y)
sample_data(phyloseq_obj_css)$thetao      <- as.numeric(sample_data(phyloseq_obj_css)$thetao)

bray_dist <- phyloseq::distance(
  phyloseq_obj_css,
  method = "bray"
)
unifrac_dist <- phyloseq::distance(
  phyloseq_obj_css,
  method = "unifrac",
  weighted = FALSE
)

coords <- data.frame(
  lon = sample_data(phyloseq_obj_css)$Longitude.y,
  lat = sample_data(phyloseq_obj_css)$Latitude.y
)

geo_dist <- distm(coords, fun = distHaversine) / 1000
geo_dist <- as.dist(geo_dist)


#distanza ambientale T
temp_dist <- dist(sample_data(phyloseq_obj_css)$thetao)

dd_df <- tibble(
  geographic_km = as.vector(geo_dist),
  delta_temp    = as.vector(temp_dist),
  bray          = as.vector(bray_dist),
  unifrac       = as.vector(unifrac_dist)
) %>%
  tidyr::drop_na()

#similaruty
dd_df <- dd_df %>%
  mutate(
    bray_sim    = 1 - bray,
    unifrac_sim = 1 - unifrac
  )
head(dd_df)
```

```{r}
#Bray–Curtis vs distanza geografica

ggplot(dd_df, aes(geographic_km, bray_sim)) +
  geom_point(alpha = 0.5, size = 2) +
  geom_smooth(method = "lm", se = TRUE, color = "black") +
  theme_bw() 
#UniFrac vs distanza geografica
ggplot(dd_df, aes(geographic_km, unifrac_sim)) +
  geom_point(alpha = 0.5, size = 2) +
  geom_smooth(method = "lm", se = TRUE, color = "black") +
  theme_bw()
#Bray–Curtis vs differenza di temperatura
ggplot(dd_df, aes(delta_temp, bray_sim)) +
  geom_point(alpha = 0.5, size = 2) +
  geom_smooth(method = "lm", se = TRUE, color = "black") +
  theme_bw() 
#UniFrac vs differenza di temperatura
ggplot(dd_df, aes(delta_temp, unifrac_sim)) +
  geom_point(alpha = 0.5, size = 2) +
  geom_smooth(method = "lm", se = TRUE, color = "black") +
  theme_bw() 
#km vs temp
ggplot(dd_df, aes(delta_temp, geographic_km)) +
  geom_point(alpha = 0.5, size = 2) +
  geom_smooth(method = "lm", se = TRUE, color = "black") +
  theme_bw()

```

tutti insieme z scored

```{r}
dd_long <- dd_df %>%
  dplyr::select(
    geographic_km,
    delta_temp,
    bray_sim,
    unifrac_sim
  ) %>%
  tidyr::pivot_longer(
    cols = c(bray_sim, unifrac_sim),
    names_to = "metric",
    values_to = "similarity"
  ) %>%
  tidyr::pivot_longer(
    cols = c(geographic_km, delta_temp),
    names_to = "distance_type",
    values_to = "distance"
  ) %>%
  dplyr::mutate(
    relation = paste(metric, distance_type, sep = " vs ")
  ) %>%
  dplyr::group_by(relation) %>%
  dplyr::mutate(
    distance_z = scale(distance)[,1],
    similarity_z = scale(similarity)[,1]
  ) %>%
  ungroup()



ggplot(dd_long, aes(distance_z, similarity_z)) +
  geom_point(alpha = 0.4, size = 1.5) +
  geom_smooth(method = "lm", se = TRUE, color = "black") +
  facet_wrap(~ relation, scales = "free") +
  labs(
    x = "Normalized distance (z-score)",
    y = "Normalized community similarity (z-score)"
  ) +
  theme_bw() +
  theme(
    strip.background = element_rect(fill = "grey90"),
    strip.text = element_text(face = "bold")
  )
```

stesso plot
```{r}
dd_plot <- dd_df %>%
  dplyr::select(
    geographic_km,
    delta_temp,
    bray_sim,
    unifrac_sim
  ) %>%
  tidyr::pivot_longer(
    cols = c(bray_sim, unifrac_sim),
    names_to = "metric",
    values_to = "similarity"
  ) %>%
  tidyr::pivot_longer(
    cols = c(geographic_km, delta_temp),
    names_to = "driver",
    values_to = "distance"
  ) %>%
  dplyr::mutate(
    metric = dplyr::recode(metric,
                           bray_sim = "Bray–Curtis",
                           unifrac_sim = "UniFrac"),
    driver = dplyr::recode(driver,
                           geographic_km = "Geographic distance",
                           delta_temp   = "Temperature difference"),
    group_id = paste(metric, driver, sep = "_")
  ) %>%
  dplyr::group_by(group_id) %>%
  dplyr::mutate(
    distance_z   = scale(distance)[,1],
    similarity_z = scale(similarity)[,1]
  ) %>%
  dplyr::ungroup()
ggplot(dd_plot,
       aes(x = distance_z,
           y = similarity_z,
           color = metric,
           linetype = driver)) +
  
  geom_smooth(method = "lm", se = TRUE, linewidth = 1) +
  
  labs(
    x = "Normalized distance (z-score)",
    y = "Normalized community similarity (z-score)",
    color = "Similarity metric",
    linetype = "Driver"
  ) +
  
  scale_linetype_manual(
    values = c(
      "Geographic distance" = "solid",
      "Temperature difference" = "dashed"
    )
  ) +
  
  theme_bw() +
  theme(
    legend.position = "right",
    legend.box = "vertical",
    panel.grid.minor = element_blank()
  )
```



statistics
```{r}

bray_output <- capture.output(
  pander(summary(lm(bray ~ geographic_km, data = dd_df)))
)
writeLines(bray_output, paste0(path, "DISTDECAY_statistics_bray_geographic_km.txt"))

unifrac_output <- capture.output(
  pander(summary(lm(unifrac ~ geographic_km, data = dd_df)))
)
writeLines(unifrac_output, paste0(path, "DISTDECAY_statistics_unifrac_geographic_km.txt"))

```



```{r}
# Mantel test semplici
mantel(bray_dist, geo_dist, permutations = 9999)
```
Mantel statistic r: 0.7684 
     Significance: 1e-04

```{r}
mantel(unifrac_dist, geo_dist, permutations = 9999)

```
Mantel statistic r: 0.4726 
      Significance: 1e-04

```{r}
# Mantel test parziali (controllando per temperatura)
mantel.partial(bray_dist, geo_dist, temp_dist, permutations = 9999)
```

Mantel statistic r: 0.4001 
      Significance: 0.0058

```{r}
mantel.partial(unifrac_dist, geo_dist, temp_dist, permutations = 9999)

```
Mantel statistic r: 0.2341 
      Significance: 0.0551 

```{r}
env_data <- sample_data(phyloseq_obj_css)[, c(
  "thetao", "so", "chl", "o2", "no3", "po4", "si"
)]

env_dist <- dist(scale(env_data))

# Mantel ambientale
mantel(bray_dist, env_dist,permutations = 9999)
mantel(unifrac_dist, env_dist,permutations = 9999)
```
```{r}
ggplot(env_data, aes(x = so, y = thetao)) +
  geom_point(alpha = 0.6, color = "blue") +
  labs(x = "Salinità (psu)", 
       y = "Temperatura potenziale (°C)",
       title = "Diagramma T-S") +
  theme_minimal() +
  # Aggiungi linee di densità dell'acqua (opzionale)
  geom_density_2d(color = "black", alpha = 0.5)
```



```{r}
env_data_completo <- data.frame(
  # I tuoi dati ambientali
  env_data,
  # Aggiungi le variabili categorical dal sample_data
  Zona = sample_data(phyloseq_obj_css)$Zona,
  Nominativo = sample_data(phyloseq_obj_css)$Nominativo.campione.
)



ggplot(env_data_completo, aes(x = so, y = thetao)) +
  geom_point(aes(color = Zona), size = 3) +
  geom_text_repel(aes(label = Nominativo), 
                  size = 2.5, 
                  box.padding = 0.3,
                  max.overlaps = 10) +
  geom_density_2d(color = "gray50", alpha = 0.5, linetype = "dotted") +
  labs(x = "Salinità (psu)", 
       y = "Temperatura potenziale (°C)",
       title = "Diagramma T-S con densità e Zona",
       color = "Zona") +
  scale_color_brewer(palette = "Set1") +  # o usa scale_color_viridis_d()
  theme_bw()


```

```{r}
env_data_completo <- data.frame(
  # Dati ambientali
  env_data,
  # Variabili categorical e coordinate
  Zona = sample_data(phyloseq_obj_css)$Zona,
  Nominativo = sample_data(phyloseq_obj_css)$Nominativo.campione.,
  Latitude = sample_data(phyloseq_obj_css)$Latitude.x,  # usando .x
  Longitude = sample_data(phyloseq_obj_css)$Longitude.x,  # usando .x
  # Componenti della corrente
  uo = sample_data(phyloseq_obj_css)$uo,
  vo = sample_data(phyloseq_obj_css)$vo
)

# Calcola velocità della corrente
env_data_completo$curr_speed <- with(env_data_completo, sqrt(uo^2 + vo^2))

# Rimuovi righe con coordinate NA
env_data_completo <- env_data_completo %>%
  filter(!is.na(Latitude), !is.na(Longitude))
```

```{r}
ggplot(env_data_completo, aes(x = so, y = thetao)) +
  # Punti base
  geom_point(aes(color = Zona), size = 3, alpha = 0.6) +
  # Frecce della corrente (scalate per visualizzazione)
  geom_segment(aes(xend = so + uo/10,  # fattore di scala per visualizzare
                   yend = thetao + vo/10,
                   color = Zona),
               arrow = arrow(length = unit(0.1, "cm")),
               alpha = 0.8) +
  geom_text_repel(aes(label = Nominativo), 
                  size = 2.5, 
                  box.padding = 0.3,
                  max.overlaps = 15) +
  scale_color_brewer(palette = "Set1", name = "Zona") +
  labs(x = "Salinità (psu)", 
       y = "Temperatura potenziale (°C)",
       title = "Diagramma T-S con vettori di corrente",
       subtitle = "Le frecce indicano direzione e intensità (scalata)") +
  theme_bw()
```

```{r}
# Velocità risultante delle correnti
curr_speed <- with(sample_data(phyloseq_obj_css),
                   sqrt(uo^2 + vo^2))

curr_dist <- dist(curr_speed)

# Test: correnti vs comunità
mantel(bray_dist, curr_dist,permutations = 9999)
mantel(unifrac_dist, curr_dist,permutations = 9999)
```
















